Universals of word order reflect optimization of grammars for efficient communication

### 2019-11-11 writeup/3-paper-names.tex
code

@@

\title{Universals of word order result from optimization of grammars for efficient communication}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
%\author[a,1,2]{Michael Hahn}
%\author[a]{Daniel Jurafsky}
%\author[b]{Richard Futrell}

%\affil[a]{Stanford University}
%\affil[b]{University of California, Irvine}

@@

% Please add here a significance statement to explain the relevance of your work
% limit 120 words 
\significancestatement{
What explains the universal properties of human languages? We present evidence that a major subset of these properties can be explained by viewing languages as codes for efficient communication among agents with highly generic cognitive constraints. In doing so, we provide the first full formalization and computational implementation of ideas which have been stated informally in the functional linguistics literature for decades. The success of this approach suggests a new way to conceptualize human language in quantitative and computational work, as an information-theoretic code dynamically shaped by communicative and cognitive pressures. Our results argue against the idea that the distinctive properties of human language result from essentially arbitrary genetic constraints.% (``universal grammar'').
}

% Please include corresponding author, author contribution and author declaration information
%\authorcontributions{MH and RF designed research. MH implemented models and experiments. MH and RF wrote the paper. MH, DJ, and RF provided comments on the paper.}
\authordeclaration{The authors declare no conflict of interest.}
%\equalauthors{\textsuperscript{1}A.O.(Author One) and A.T. (Author Two) contributed equally to this work (remove if not applicable).}
\correspondingauthor{\textsuperscript{2}To whom correspondence should be addressed. E-mail: mhahn2@stanford.edu}

@@

\begin{abstract}
%	TODO rewrite
The universal properties of human languages have been the subject of intense study across disciplines.
	We report novel computational and corpus-based evidence for the hypothesis that a prominent subset of these universal properties---those related to word order---result from a process of optimization for efficient communication among humans.
	We computationally formalize functional theories of language efficiencies using information theory and state-of-the-art machine learning techniques.
Applying these methods to data from 51 languages, we show that evolution towards efficiency results in word order patterns that predict a large subset of the major word order correlations reported in the linguistic typological literature.
%	TODO explain
	%We develop a probabilistic, differentiable model of word order grammars: the means by which different languages convert underlying hierarchical structures into strings of words.
	%We show how the parameters of a word order grammar can be optimized for efficiency of information transfer, quantified as a tradeoff between incremental predictability and mutual information with latent tree structures.
%	Applying these grammars to tree structures found in dependency corpora from 51 languages, we show that optimizing the grammar parameters for efficiency results in word order patterns that predict a large subset of the major word order correlations reported in the linguistic typological literature, and subsumes the predictions of previous heuristic theories such as dependency length minimization.
\end{abstract}

@@

\begin{document}
% TODO mention the preregistration somewhere
\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}

%Ray Jackendoff. 2002. Foundations of Language:
%Brain, Meaning, Grammar, Evolution. Oxford
%University Press, Oxford, UK

%William Croft and Alan Cruse. 2004. Cognitive
%Linguistics. Cambridge University Press, Cam-
%bridge, UK.
% --> emphasize language use

%Adele Goldberg. 2005. Constructions at work:
%The nature of generalization in language. Ox-
%ford University Press, Oxford, UK.
% --> emphasize usage-based grammar
% we are all born with the same basic conceptual apparatus, with the
%same basic communicative demands, and we all live in the physical world with
%forces of gravity, bodies, and night and day (cf. also LakoV 1987; Webelhuth and
%Ackerman 1998). Certain semantic/pragmatic functions are so relevant and
%useful that they can be found in language after language (Croft 2001).
%Cross-linguistic generalizations that relate form and function are explained
%by appeal to general cognitive constraints together with the functions of the
%constructions involved. Constructionists turn to grammar-external explan-
%ations such as universal functional pressures, iconic principles, and processing
%and learning constraints to explain such empirically observable cross-linguis-
%tic generalizations.
%For example, the fact
%that that most languages seem to have noun and verb categories may be
%explained by the existence of corresponding basic semantic categories (Baker
%2004). Hauser, Chomsky, and Fitch go so far as to suggest that the only innate
%ability speciWc to language that may be required are the recursive interfaces to
%phonology and semantics, and they allow that even these may turn out not to
%be speciWc to language (Hauser, Chomsky, and Fitch 2002).

% Haspelmath 1999 talks about types of explanations

%Joan Bresnan, Ash Asudeh, Ida Toivonen, and
%Stephen Wechsler. 2016. Lexical-Functional
%Syntax, 2nd ed. Blackwell, Malden, MA.

%Andrew Radford. 2006. Minimalist syntax re-
%visited. http://www.public.asu.edu/
%~gelderen/Radford2009.pdf.

%Givon

% TODO optimality theory and syntax citation?
%Blutner R. (2000) Some aspects of optimality in natural language interpretation. Journal of Semantics 17: 189–216

% ,lin2017critical
\dropcap{U}nderstanding and explaining the variation between human languages is the central  goal of Linguistics. % \cite{greenberg1963universals,chomsky1993lectures,givon2009genesis, ...}.
% behaghel1909beziehungen,
% \cite{behaghel1909beziehungen,zipf1949human,greenberg1963universals,lin2017critical} \cite{hawkins2007processing,bender2009linguistically,bender2013linguistic}
%An explanation for the universal properties of language would enable a deeper scientific understanding of what human language is and how to model it, with applications in psychology and natural language processing \cite{hawkins2007processing,bender2009linguistically,bender2013linguistic}.
% TODO mention universals here
Across theoretical paradigms, linguists have hypothesized that language is shaped by efficiency in computation~\cite{chomsky2005three,hauser2002faculty, berwick1984grammatical,hawkins1994performance} and communication~\cite{zipf1949human, Croft:Cruse:2004, Goldberg:2005, pinker1990natural, smith2013linguistic}.
%Role of general computational constraints ~\cite{hauser2002faculty} % berwick2016only
In this work we examine this question from a computational perspective, demonstrating a fully formalized and implemented framework in which a large class of language universals can be explained through the computational optimization of grammars for efficiency. % information-theoretic 
Specifically, we show that efficiency optimization explains  linguistic universals related to word order first documented by Greenberg \cite{greenberg1963universals}. 
% TODO change wording so it's clear that this is a very major thing

% Things TODO
% - make clear that efficiency is something both functional and generative linguists have been appealing to
% - at some point address head direction idea
%TODO at some point address head direction idea
%    - Vennemann 1972, 1974ab, 1976
%    -Hawkins 1980 limitations of head direction, also Hawkins 1982

%foreshadow functional explanations?
%
%- these linguistic patterns reflect general principles of communication
%
%- big idea has been around for a long time
%
%- here for the first time direct computational test
%
%- we can predict broad patterns, and make language-specific predictions based on these general principles 
%

% TODO add a cogsci person to citation chain at the beginning

%\dropcap{F}or decades, researchers in fields ranging from philology to cognitive science to statistical physics have been involved in documenting and trying to explain the universal syntactic and statistical properties of human language \cite{behaghel1909beziehungen,zipf1949human,greenberg1963universals,lin2017critical}. An explanation for the universal properties of language would enable a deeper scientific understanding of what human language is and how to model it, with applications in psychology and natural language processing \cite{hawkins2007processing,bender2009linguistically,bender2013linguistic}.

%Berwick and Chomsky ‘gene rative process  is optimal  ’, ‘efficient computa  tion ’(71), ‘this newly  emerged  computationalsystem forthought  ...ispe rfe ct, in sofar asSMT is corr ect ’(80).  ‘op -tim al’or ‘effici ent ’or ‘per fect ’.

%Chomsky Three Factors, 2003:
% --> emphasizes 'computational efficiency'

%\textbf{TODO maybe should be more general in the beginning, as in the first version. Also make sure the reader doesn't think 'it's just headedness, why bother'.}

%We aim to explain the linguistic universals related to word order first documented by Greenberg \cite{greenberg1963universals}. 
Human languages vary in the order in which they express information.
Consider Figure~\ref{fig:arabic-japanese-simple}, showing a sentence in Arabic (top) and Japanese (bottom), both translating to `I wrote a letter to a friend.'
Both sentences contain a verb meaning `wrote', a noun expressing  the object `letter', and an adpositional phrase translating to `to a friend'.
%In linguistic terminology, `letter' is the object of the verb, whereas the phrases translating as `to a friend' are known as adpositional phrases.
Yet, the order of these words are entirely different in the two languages:
%However, the languages differ in the order in which these words and phrases are arranged: 
The verb stands at the beginning in Arabic, and at the end in Japanese.
Arabic expresses `to' by a \emph{preposition} (\emph{preceding} the noun `friend'); Japanese uses a \emph{postposition} (\emph{following} it). % and a postposition in Japanese, following `friend'.

This variation reflects a deep and stable regularity:
While languages ordering the objects before (Japanese) or after (Arabic) the verb are approximately equally common around the world,
this is strongly correlated with the occurrence of pre- or postpositions (Figure~\ref{fig:corr-table}, top):
Languages ordering their objects the way Japanese does, have postpositions; languages ordering them as Arabic does have prepositions.

\begin{figure}
    \centering
    \includegraphics[width=0.4\textwidth]{figures/corr-full.png}
\caption{ The concept of a word order correlation: Languages can order the object after (Arabic) or before (Japanese) the verb, and have prepositions (Arabic) or postpositions (Japanese). For each combination, we indicate how many languages satisfy it, as documented in the World Atlas of Language Structure~\cite{wals}. Combinations on the diagonal are vastly more common than off-diagonal ones. % A sentence in Arabic (top) and Japanese (bottom), translating to `I wrote a letter to a friend.' Note the reversal of word order: Arabic has verb-object order and prepositions, while Japanese has object-verb order and postpositions.
	}\label{fig:arabic-japanese-simple}	\label{fig:corr-table}
\end{figure}

This generalization falls in a group of language universals that were originally documented by Greenberg \cite{greenberg1963universals}, known as \key{word order correlations}:
These describe correlations between the relative positions of different types of expressions across languages.
The example above documents that the position of the object (`letter') relative to the verb is \key{correlated} with the position of the adposition (`to'). %adpositional phrases (`to a friend') relative to the verb.
Greenberg also found that the order between verb and object is correlated with other aspects of a language's word order (Table~\ref{table:corr-dryer}), such the order of verb and adpositional phrase (`wrote -- to friend' in Arabic vs `friend to -- wrote' in Japanese), and that of noun and genitive (`book -- of friend' in Arabic, `friend of -- book' in Japanese).

@@

Explaining these patterns has been an important aim of linguistic research since Greenberg's seminal study~\cite{lehmann1973structural, vennemann1974theoretical,jackendoff1977x,frazier1985syntactic,chomsky1988language, dryer1992greenbergian, hawkins1994performance}.

A prominent line of research has argued that language universals arise for \key{functional} reasons: that is, because they make human communication and language processing maximally efficient, and regularities across languages hold because these efficiency constraints are rooted in general principles of communication and cognition (e.g., \cite{gabelentz1901sprachwissenschaft,zipf1949human,hockett1960origin,givon1991markedness,hawkins1994performance,hawkins2004efficiency,hawkins2014crosslinguistic,croft2001functional,haspelmath2008parametric,jaeger2011language}).
Under this view, the various human languages represent multiple solutions to the problem of efficient information transfer given human cognitive constraints.

Zipf \cite{zipf1949human} argued that language optimizes a tradeoff between the Forces of \emph{Unification} and \emph{Diversification}.
The Force of Unification is a pressure to reduce the complexity of the language to make production and processing as easy as possible.
Conversely, the Force of Diversification favors languages that provide different utterances for different meanings, so that the listener can unambiguously identify the meaning from the utterance.
These two forces act in opposing directions:
Producing and processing simple utterances incurs little cost, but more complex and diverse utterances are required in order to provide higher informativity.
The idea that many properties of language arise from the tension between these two pressures has a long and fruitful history in theoretical studies of language~\cite{gabelentz1901sprachwissenschaft,horn1984toward, lindblom1990explaining, schwartz1997dispersion, hawkins2004efficiency}.

Recent work has used an information-theoretic formalization to computationally test these ideas in various domains of language, showing that they predict
both basic statistical properties of languages~\cite{ferreri2003least} and language development~\cite{smith2013linguistic},
and sophisticated aspects of language, such as pragmatic inference \cite{frank2012predicting}, and the distribution of color words~\cite{zaslavsky2018efficient} and kinship categories~\cite{kemp2012kinship} across many languages.

While it has been suggested that these ideas should also apply to grammar~\cite{hawkins2004efficiency}, testing these accounts is more difficult in the domain of grammar, as this requires both large amounts of data representative of language use across languages and cultures, and computational methods for estimating the efficiency of entire languages.
In this work, we address these challenges by combining large-scale text data from 51 languages with state-of-the-art machine learning techiques to estimate the communicative efficiency of grammar.
We apply this approach to the problem of explaining Greenberg's word order correlation universals.
We first show that the word order of languages has evolved to optimize efficiency.
We then show that efficiency optimization accounts for the Greenbergian word order correlations, and makes fine-grained per-language predictions about individual syntactic patterns.
Our results show how general properties of efficient communication gives rise to the universal word order properties of human language.

\begin{table}
	\begin{center}
%		\setlength\tabcolsep{1.5pt}
\begin{tabular}{|c|ll|ll|}
	\hline
	%&
	&	\multicolumn{2}{c|}{\textbf{Arabic} (English, ...)}   &        \multicolumn{2}{c|}{\textbf{Japanese} (Hindi, ...)} \\ \hline\hline
	& \multicolumn{2}{c}{Correlates with...} & \multicolumn{2}{|c|}{Correlates with...}  \\
	&	Verb & Object     & Object & Verb    \\ \hline\hline
	& kataba & rasaa'il 	&tegami-o & kaita \\
	&	\emph{wrote} & \emph{letters} & 	\emph{letter} & \emph{wrote} \\ \hline 
	\multirow{2}{*}{\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {1}}}}	&	li    &    sadiiq       &	tomodachi & ni \\
	&		\emph{to}            & \emph{a friend} &		\emph{friend} & \emph{to} \\ \hline
	\multirow{2}{*}{\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {2}}}}	&kaana    &    sadiiq         &	tomodachi & datta \\
	&	\emph{was}        & \emph{a friend} 	&\emph{friend} & \emph{was} \\ \hline
	\multirow{2}{*}{\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {3}}}}	&sawfa    &    yaktub       & 	kak & udesho \\
	&	\emph{will}          & \emph{write}  &    	\emph{write} & \emph{will} \\ \hline
	\multirow{2}{*}{\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {4}}}}	& sadiiq  &    John    & 	John no & tomodachi \\
	&	\emph{friend} &  \emph{of John}  &	\emph{John of} & \emph{friend} \\ \hline
	\multirow{2}{*}{\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {5}}}}	&kutub    &    taqra'uha       & 	anata-ga yonda & hon \\
	&	\emph{books} & \emph{that you read}  &	\emph{that you read} & \emph{book} \\ \hline
	\multirow{2}{*}{\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {6}}}}	&'an    &    tusil        & 	toochaku suru & koto \\
	&	\emph{that} & \emph{she arrives}  &	\emph{has.arrived} & \emph{that} \\ \hline
	\multirow{2}{*}{	\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {7}}}}	&dhahabt	    &    'ila lmadrasa     & 	gakkoo ni & ittekimashita \\
	&	\emph{went} & \emph{to school}  &	\emph{school to} & \emph{went} \\ \hline
	\multirow{2}{*}{\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {8}}}}	&'uriid    &    'an 'ughaadir         & 	yom & itai \\
	& \emph{wants}   &  \emph{to leave}  &	\emph{to leave} & \emph{want} \\ \hline \hline
\end{tabular}
	\end{center}
	\caption{Greenberg's word order correlations, exemplified by Arabic (left) and Japanese (right) examples: Across the world, the orders of different constituents are strikingly correlated with that of verb and object. See Materials and Methods.}\label{table:corr-dryer}
\end{table}

\section{Formalizing Efficiency in Word Order}

\begin{figure}
    \centering
    \includegraphics[width=0.45\textwidth]{figures/communication-two-langs.png}
	\caption{Grammars define communication channels: A speaker (left) expresses a dependency structure into a string of words forming a sentence. The order of words is determined by the grammar of the language. A listener recovers a dependency structure. Communication is successful if the listener recovers the correct depenency structure.}
	\label{fig:comm}
\end{figure}

% Given a set of dependency structures, grammars define a set of sentences. Each grammar defines a consistent ordering for the different kinds of syntactic relations, e.g., 
\begin{figure}
    \centering
    \includegraphics[width=0.45\textwidth]{figures/grammars-langs.png}
	\caption{Grammars define consistent ordering rules for dependency structures. Here, Grammars 1 and 2 order the object after the verb, Grammars 3 and 4 order the object before the verb. Grammars 1 and 3 conform to the Greenberg correlations and are common around the world; Grammars 2 and 4 are rare or impossible.}
	\label{fig:grammars}
\end{figure}

We model the process where a speaker communicates a message to a listener (Figure~\ref{fig:comm}) in Shannon's framework of information theory.
In this model, a transmitter encodes a message into a signal. The receiver decodes the signal, attempting to reconstruct the original message.
Applying this model to word order, we take the message to consist of people, objects, events, and a set of syntactic and semantic relations that hold between them~\cite{fodor1975language,jackendoff1983semantics,pinker1990natural}.
Following a long tradition in formal and computational liguistics, we formalize these in the format of dependency trees~\cite{hays1964dependency,melcuk1988dependency,corbett1993heads,tesniere2015elements}:
This linguistic formalism draws directed arcs between syntactically related words, annotated with syntactic relations.
Although there are some differences in how different formalisms would draw trees for certain sentences, there is broad enough agreement about dependency trees that it has been possible to develop large-scale dependency-annotated corpora of text from dozens of languages \cite{nivre2017universal}.

%TODO also need to say these are syntax \& principle of compositionality

>>>>

%https://tex.stackexchange.com/questions/219676/how-to-change-spacing-before-and-after-paragraph-command
%\usepackage{xpatch}

@@

\renewcommand{\baselinestretch}{0.99}
\addtolength{\textfloatsep}{-0.18in}
\setlength{\abovecaptionskip}{5pt plus 1pt minus 0pt} % Chosen fairly arbitrarily
%\setlength{\belowcaptionskip}{-5pt}
%https://tex.stackexchange.com/questions/23313/how-can-i-reduce-padding-after-figure

\title{Universals of word order reflect optimization of grammars for efficient communication}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a,1]{Michael Hahn}
\author[a]{Dan Jurafsky}
\author[b]{Richard Futrell}

\affil[a]{Stanford University}
\affil[b]{University of California, Irvine}

@@

% Please add here a significance statement to explain the relevance of your work
% limit 120 words 
\significancestatement{
Human languages share many grammatical properties.
We show that some of these properties 
can be explained by the need for languages to offer
efficient communication between humans given our cognitive constraints. Grammars of languages seem to find a balance between two communicative pressures: to be simple enough to allow the speaker to easily produce sentences, but complex enough to be unambiguous to the hearer, and this balance explains well-known word order generalizations across our sample of 51 varied languages.   Our results offer new quantitative and computational evidence that language structure is dynamically shaped by communicative and cognitive pressures.}

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{MH, DJ, and RF designed research. MH implemented models and experiments. MH, DJ, and RF wrote the paper.}
\authordeclaration{The authors declare no conflict of interest.}
\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: mhahn2@stanford.edu}

@@

\begin{abstract}
The universal properties of human languages have been the subject
of intense study across the language sciences.
We report novel computational and corpus evidence for
the hypothesis that a prominent subset of these universal
properties---those related to word order---result from a
process of optimization for efficient communication among humans,
trading off the need to reduce complexity with the need to reduce ambiguity.
We formalize these two pressures with information-theoretic and neural network models
of complexity and ambiguity, and simulate grammars with optimized
word order parameters on large-scale data from 51 languages. 
Evolution of grammars towards efficiency results in word order patterns that
predict a large subset of the major word order correlations 
across languages.
\end{abstract}

@@

\begin{document}
\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}

\dropcap{U}nderstanding what is universal and what varies across human languages is a central goal of linguistics.
Across theoretical paradigms, linguists have hypothesized that language is shaped by efficiency in computation~\cite{chomsky2005three,hauser2002faculty, berwick1984grammatical,hawkins1994performance} and communication~\cite{zipf1949human, Croft:Cruse:2004, Goldberg:2005, kirby2015compression, pinker1990natural, nowak1999evolution}. %nowak2001towards, nowak2002computational
But formalizing how these pressures explain specific grammatical universals has proved difficult.
Here we pair new computational models that measure the communicative efficiency of grammars with a new simulation framework for finding optimal grammars, and show that the most efficient grammars also exhibit a large class of language universals. % information-theoretic

The language universals we study are the well-known Greenberg universals of word order \cite{greenberg1963universals}. 
Human languages vary in the order in which they express information.
Consider Figure~\ref{fig:arabic-japanese-simple}, showing a sentence in Arabic (top) and Japanese (bottom), both translating to `I wrote a letter to a friend.'
Both sentences contain a verb meaning `wrote', a noun expressing  the object `letter', and a phrase translating to `to a friend'.
Yet the order of these words are entirely different in the two languages:
the verb stands at the beginning in Arabic, and at the end in Japanese.
Arabic expresses `to' by a \emph{preposition} (\emph{preceding} the noun `friend'); Japanese uses a \emph{postposition} (\emph{following} it).

Yet this variation reflects a deep and stable regularity:
while languages ordering the objects before (Japanese) or after (Arabic) the verb are approximately equally common around the world,
this is strongly correlated with the occurrence of pre- or postpositions (Figure~\ref{fig:corr-table}, top):
Languages ordering their objects the way Japanese does have postpositions; languages ordering them as Arabic does have prepositions.

\begin{figure}
    \centering
    \includegraphics[width=0.4\textwidth]{figure1.pdf}
\caption{ One word order correlation: Languages can order the object after (Arabic) or before (Japanese) the verb, and have prepositions (Arabic) or postpositions (Japanese). For each combination, we indicate how many languages satisfy it, as documented in the World Atlas of Language Structures~\cite{wals}. Combinations on the diagonal are vastly more common than off-diagonal ones. 
	}\label{fig:arabic-japanese-simple}	\label{fig:corr-table}
\end{figure}

This generalization lies in a group of language universals originally documented by Greenberg \cite{greenberg1963universals}, known as \key{word order correlations}.
These describe correlations between the relative positions of different types of expressions across languages.
The example above documents that the position of the object (`letter') relative to the verb is \key{correlated} with the position of the adposition (`to'). %adpositional phrases (`to a friend') relative to the verb.
Greenberg also found that the order of verb and object is correlated with other aspects of a language's word order (Table~\ref{table:corr-dryer}), such as the order of verb and adpositional phrase (`wrote -- to friend' in Arabic vs `friend to -- wrote' in Japanese), and that of noun and genitive (`book -- of friend' in Arabic, `friend of -- book' in Japanese).

\begin{table}[ht]
	\begin{center}
\begin{tabular}{|c|ll|ll|}
	\hline
	&	\multicolumn{2}{c|}{\textbf{Arabic} (English, ...)}   &        \multicolumn{2}{c|}{\textbf{Japanese} (Turkish, ...)} \\ \hline\hline
	& \multicolumn{2}{c|}{Correlates with...} & \multicolumn{2}{c|}{Correlates with...}  \\
	&	Verb & Object     & Object & Verb    \\ \hline\hline
	& kataba & ris{\= a}la 	&tegami-o & kaita \\
	&	\emph{wrote} & \emph{letter} & 	\emph{letter} & \emph{wrote} \\ \hline 
	\multirow{2}{*}{\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {1}}}}	&	li    &    {\d s}ad{\= i}q       &	tomodachi & ni \\ % G3, G4
	&		\emph{to}            & \emph{a friend} &		\emph{friend} & \emph{to} \\ \hline 
	\multirow{2}{*}{\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {2}}}}	&k{\= a}na    &    {\d s}ad{\= i}q         &	tomodachi & datta \\ % not in Greenberg
	&	\emph{was}        & \emph{a friend} 	&\emph{friend} & \emph{was} \\ \hline
	\multirow{2}{*}{\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {3}}}}	&sawfa    &    yaktub       & 	kak- & -udesho \\ % G16, G13 TODO this should only be inflected auxiliaries, replace Arabic
	&	\emph{will}          & \emph{write}  &    	\emph{write} & \emph{will} \\ \hline
	\multirow{2}{*}{\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {4}}}}	& {\d s}ad{\= i}q  &    John    & 	John no & tomodachi \\ % G2, G23
	&	\emph{friend} &  \emph{of John}  &	\emph{John of} & \emph{friend} \\ \hline
	\multirow{2}{*}{\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {5}}}}	&kutub    &    taqra'uh{\= a}       & 	anata-ga yonda & hon \\ % G24
	&	\emph{books} & \emph{that you read}  &	\emph{that you read} & \emph{book} \\ \hline
	\multirow{2}{*}{\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {6}}}}	&'an    &    tu{\d s}il        & 	toochaku suru & koto \\ % 
	&	\emph{that} & \emph{she arrives}  &	\emph{arrives} & \emph{that} \\ \hline
	\multirow{2}{*}{	\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {7}}}}	&dhahabt	    &    'il{\= a} lmadrasa     & 	gakkoo ni & itta \\ % G22
	&	\emph{went} & \emph{to school}  &	\emph{school to} & \emph{went} \\ \hline
	\multirow{2}{*}{\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {8}}}}	&'ur{\= i}d    &    'an 'ugh{\= a}dir         & 	ik- & -itai \\ % G15
	& \emph{wants}   &  \emph{to leave}  &	\emph{to go} & \emph{want} \\ \hline 
\end{tabular}
	\end{center}
	\caption{Greenberg's word order correlations, exemplified by Arabic (left) and Japanese (right) examples: Across the world, the orders of different constituents are strikingly correlated with that of verb and object.
	Selection is based on a more recent typological study by Dryer~\cite{dryer1992greenbergian}, restricted to those correlations that are annotated in available corpus data. See SI Section S1 for more on Greenberg correlations.
	}\label{table:corr-dryer}
\end{table}

@@

Explaining these patterns has been an important aim of linguistic research since Greenberg's seminal study~\cite{lehmann1973structural, jackendoff1977x,frazier1985syntactic,chomsky1993theory, kayne2003antisymmetry, baker2008macroparameter, dryer1992greenbergian, hawkins1994performance}. % cinque2005deriving, vennemann1974theoretical,
Prominent among this research is the argument that
language universals arise for \key{functional} reasons: that is, because they make human communication and language processing maximally efficient, and regularities across languages hold because these efficiency constraints are rooted in general principles of communication and cognition (e.g., \cite{gabelentz1901sprachwissenschaft,zipf1949human,hockett1960origin,pinker1990natural,givon1991markedness,hawkins1994performance,hawkins2004efficiency,croft2001functional,haspelmath2008parametric,jaeger2011language,kirby2015compression}).
Under this view, the various human languages represent multiple solutions to the problem of efficient information transfer given human cognitive constraints.

In an early and  influential functional framework,
Zipf \cite{zipf1949human} argued that language optimizes a tradeoff between two pressures: to reduce complexity and to reduce ambiguity.
What Zipf called the `Force of Unification' is a pressure to reduce the complexity of the language by reducing the number of distinctions made in the language, in order to make production and processing as easy as possible.
The countervailing `Force of Diversification' favors languages that provide different utterances for different meanings, so that the listener can unambiguously identify the meaning from the utterance.
These two forces act in opposing directions:
Producing and processing simple utterances incurs little cost, but more complex and diverse utterances are required to provide enough information.
The idea that many properties of language arise from the tension between these two pressures has a long and fruitful history in linguistics~\cite{gabelentz1901sprachwissenschaft,horn1984toward,lindblom1990explaining,hawkins2004efficiency,haspelmath2008functional}. % , wedel2013high, schwartz1997dispersion

Recent work has drawn on information theory to computationally test this ``dual pressures'' idea in various domains of language, showing that it predicts
both basic statistical properties of languages~\cite{ferreri2003least,torre2019physical} and language evolution~\cite{kirby2015compression},
and sophisticated aspects of language, such as pragmatic inference \cite{frank2012predicting}, and the distribution of color words~\cite{zaslavsky2018efficient} and kinship categories~\cite{kemp2012kinship} across many languages.
While it has been suggested that the dual pressure should also apply to grammar~\cite{hawkins2004efficiency}, testing these accounts is more difficult, as this requires large amounts of data representative of language use across languages,  computational methods for estimating the efficiency of entire languages, and a simulation methodology for comparing different possible grammars.

In this work, we address these challenges by combining large-scale text data from 51 languages with machine learning techniques to estimate both aspects of the communicative efficiency of grammar: complexity and ambiguity.
We use machine learning models based on neural networks to model the evolution of grammars towards efficiency.
We apply this approach to the problem of explaining Greenberg's word order correlation universals.

In Study 1 we compare the word order of actual grammars of 51 languages with alternative ``counterfactual'' grammars parameterized by different word orders. We use our model to measure the communicative efficiency of each possible grammar, showing  that the grammars of real languages are more efficient than alternative grammars. The fact that real grammars lie at the Pareto frontier of the efficiency space of possible grammars suggests that the word order of languages has evolved to optimize communicative efficiency.

In Study 2 we test whether efficiency optimization accounts for the Greenbergian word order correlations.  For each of the 51 languages, we create hypothetical grammars optimized for efficiency.  We then test statistically whether these optimized grammars exhibit the Greenberg correlations, using a Bayesian mixed-effects logistic regression to control for language and language family.  Efficiency optimization indeed predicts all eight Greenberg correlations.
Our results show that general properties of efficient communication can give rise to these universal word order properties of human language.

\section*{Grammars and Grammar Data}

Following a long tradition in theoretical and computational linguistics, we formalize  the grammatical structure of languages using dependency trees~\cite{hays1964dependency,hudson1984word,melcuk1988dependency,corbett1993heads,tesniere2015elements}.
This linguistic formalism represents grammatical dependencies as directed arcs between syntactically related words, annotated with grammatical relations like subject or object (Figure~\ref{fig:sent-dep}).
While syntactic formalisms vary, the dependency grammar community has an agreed representation format  which has been used to  annotate corpora of text from dozens of languages \cite{ud2.1}, and there are 
computational methods for deriving such representations from other standard linguistic formalisms \cite{boston2009dependency}.

\begin{figure}[ht]
    \centering
{
\centering{
\begin{dependency}[theme = simple]
   \begin{deptext}[column sep=1em]
          She \& met \& friends  \\
   \end{deptext}
   \depedge{2}{1}{subject}
   \depedge{2}{3}{object}
\end{dependency}
}
}
	\caption{An English sentence with annotated syntactic relations.}
	\label{fig:sent-dep}
\end{figure}

Our models require a sample of syntactic structures as actually used by speakers across different languages, for which we draw on the recent
Universal Dependencies project~\cite{ud2.1}, which has collected and created syntactic annotations for several dozens of different languages.
51 languages had sufficient data for our purposes. 
These corpora represent a typologically and genetically diverse group of languages. %  (see Figure~\ref{fig:per-lang})
We obtained a total of 11.7M words in 700K sentences annotated with syntactic structures; with a median of 117K words and 7K sentences for each individual language.

\section*{Study 1: Efficiency of Languages}
\label{sec:relative-efficiency}

We first ask whether the grammars of human languages reflect optimization for efficiency of communication.  To do this we
compare the efficiency of the actual grammars of the 51 languages from the Universal Dependencies datasets to randomly constructed baseline grammars.

The grammars of natural languages specify how the different words in a syntactic structure are ordered into a sentence, i.e., a string of words  \cite{adger2015syntax}.
This is illustrated in Figure~\ref{fig:grammars}:
We show how four different grammars order objects, adpositional phrases, and adpositions.
For instance, Grammar 1---corresponding to Arabic in Figure~\ref{fig:arabic-japanese-simple}---orders objects (`friends', `letter') after verbs and has prepositions (`to friend').
Grammar 2 orders objects after verbs but has postpositions (`friend -- to').
Grammars 3 and 4 place the object before the verb, and one of them (Grammar 3) corresponds to Japanese order.

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.48\textwidth]{figure3.pdf}
	\caption{Grammars define consistent ordering rules for syntactic structures. Here, Grammars 1 and 2 order the object after the verb, Grammars 3 and 4 order the object before the verb. Grammars 1 and 3 conform to the Greenberg correlations and are common around the world; Grammars 2 and 4 are rare or impossible.} %  \textcolor{red}{TODO bubble misleading?}
	\label{fig:grammars}
\end{figure}

Beyond the syntactic relations exemplified in Figure~\ref{fig:grammars}, human languages have further types of syntactic relations.
The Universal Dependencies project, the source of our data, defines a total of 37 syntactic relations.
We adopt a variant of the grammar model developed by Gildea and colleagues~\cite{gildea2007optimizing,gildea2010grammars,gildea2015human}: a grammar assigns a weight from $[-1,1]$ to each of these 37 syntactic relations, and orders words according to the weights assigned to their relations
(See \textit{Materials and Methods} for details).

Given a large database of sentences annotated with syntactic structures (such as those at the top of Figure~\ref{fig:grammars}), obtained from a corpus of some real language $L$, we can apply a grammar to reorder the structures in the database into a dataset of counterfactual sentences belonging to a hypothetical language defined by that grammar (Figure~\ref{fig:grammars}).
This hypothetical language has identical syntactic structures and grammatical relations as the true language $L$, but different word order.

We create baseline grammars by randomly sampling the weights for each syntactic relation.
These baseline grammars have systematic word order rules similar to natural language, but do not exhibit any correlations among the orderings of different syntactic relations.
All four grammars in Figure~\ref{fig:grammars} are equally likely under this baseline distribution.

For every one of the 51 languages, we construct 50 counterfactual baseline versions by randomly creating 50 baseline grammars and applying them to obtain counterfactual orderings for all syntactic structures that were available for that language.
